---
format:
  pdf:
    prefer-html: true
---

# Centroids of transport zones

In this section we will calculate the geometric and the weighted centroids of transport zones.

## Geometric centroids

Taking the `Municipalities_geo` data from the previous section, we will calculate the geometric centroids, using the `st_centroid()` function.

```{r geocentroid}
#| message: false
#| warning: false
library(dplyr)
library(sf)
library(mapview)

Municipalities_geo = st_read("data/Municipalities_geo.gpkg", quiet = TRUE)

Centroids_geo = st_centroid(Municipalities_geo)
```

This creates points at the geometric center of each polygon.

```{r}
#| message: false
#| fig-format: png
mapview(Centroids_geo)
mapview(Centroids_geo) + mapview(Municipalities_geo, alpha.regions = 0) # both maps, with full transparency in polygons
```

But...
is this the best way to represent the center of a transport zone?

These results may be biased by the shape of the polygons, and not represent where activities are.
Example: lakes, forests, etc.

To overcome this, we can use **weighted centroids**.

## Weighted centroids

We will weight centroids of transport zones by population and by number of buildings.

For this, we will need the census data [@INEcensus].

```{r}
#| eval: false
#| include: false
# data prep
CENSUS = st_read("D:/GIS/MQAT/original/BGRI21_LISBOA.gpkg")
CENSUS = CENSUS |> 
  select(DTMN21, SUBSECCAO, N_INDIVIDUOS, N_EDIFICIOS_CLASSICOS) |>
  rename(Mun_code = "DTMN21",
         UID = "SUBSECCAO",
         Population = "N_INDIVIDUOS",
         Buildings = "N_EDIFICIOS_CLASSICOS")
CENSUS = CENSUS |> left_join(Municipalities |> select(Mun_code, Municipality) |> distinct()) |> 
  select(Municipality, UID, Population, Buildings)
st_write(CENSUS, "data/census.gpkg")
```

```{r getcensus}
#| message: false
#| fig-format: png
Census = st_read("data/census.gpkg", quiet = TRUE)

mapview(Census |> filter(Municipality == "Lisboa"), zcol = "Population")
```

It was not that easy to estimate weighted centroids with R, as it is with GIS software.
But there is this new package [`centr`](https://ryanzomorrodi.github.io/centr) that can help us [@centr].

We need to specify the **group** we want to calculate the **mean centroids**, and the **weight variable** we want to use.

```{r}
# test
library(centr)
Centroid_pop = Census |> 
  mean_center(group = "Municipality", weight = "Population")
```

We can do the same for buildings.

```{r}
Centroid_build = Census |> 
  mean_center(group = "Municipality", weight = "Buildings")
```

## Compare centroids in a map

### Interactive map

```{r}
#| fig-format: png
mapview(Centroids_geo, col.region = "blue") +
  mapview(Centroid_pop, col.region = "red") +
  mapview(Centroid_build, col.region = "black") +
  mapview(Municipalities_geo, alpha.regions = 0) # polygon limits
```

See how the building, population and geometric centroids of Sintra are apart, from closer to Tagus, to the rural area.

### Static map

To produce the same map, using only `plot()` and `st_geometry()`, we need to make sure that the geometries have the same crs.

```{r}
#| eval: false
st_crs(Centroids_geo) # 4326 WGS84
st_crs(Centroid_pop) # 3763 Portugal TM06
```

So, we need to transform the geometries to the same crs.

```{r}
Centroid_pop = st_transform(Centroid_pop, crs = 4326)
Centroid_build = st_transform(Centroid_build, crs = 4326)
```

Now, to use `plot()` we incrementally add layers to the plot.

```{r}
#| fig.dpi: 300
#| fig-cap: "Static map of different centroids of Municipalities"
#| fig-format: png
# Plot the Municipalities_geo polygons first (with no fill)
plot(st_geometry(Municipalities_geo), col = NA, border = "black")

# Add the Centroids_geo points in blue
plot(st_geometry(Centroids_geo), col = "blue", pch = 16, add = TRUE) # add!

# Add the Centroid_pop points in red
plot(st_geometry(Centroid_pop), col = "red", pch = 16, add = TRUE)

# Add the Centroid_build points in black
plot(st_geometry(Centroid_build), col = "black", pch = 16, add = TRUE)
```

In the [next section](desire-lines.qmd) we will use these centroids to calculate the desire lines between them.
