---
eval: true
code-fold: false
engine: knitr
---

# R basics

In this chapter we will introduce to the R basics and some exercises to get familiar to how R works.

## Simple operations

-   Math operations

```{r}

# Sum
1+1

# Substraction
5-2

# Multiplication
2*2

# Division
8/2
```

```{r}
# Round the number

round(3.14)
round(3.14, 1) # The "1" indicates to round it up to 1 decimal digit.

# You can use help ?round in the console to see the description of the function. 
```

-   Basic shortpaths

```{r}
# Perform combinations

c(1, 2, 3)
c(1:3) # The ":" indicates a range between the first and second numbers. 
```

```{r}
# Create a comment with ctrl + shift + r
# Comments help you organize your code. The software will not run the comment. 
```

-   Create a table:

    A simple table with the number of trips by car, PT, walking, and cycling in a hypothetical street segment at a certain period.

```{r}

# Define the variables
modes <- c("car", "PT", "walking", "cycling") # you can use "=" or "<-"
Trips = c(200, 50, 300, 150) # uppercase letters modify

# Join in table
table_example = data.frame(modes, Trips)
```

```{r}
# Visualize the table by clicking on the "Data" in the "Environment" page. 
# Or use the following function:
View(table_example)
```

## Practical exercise

Import dataset with the number of trips between all municipalities in the Metropolitan Area of Lisbon, Portugal.

```{r}
data = readRDS("data/TRIPSmode_mun.Rds")
```

Take a first look at the data

```{r}
# Summary statistics
summary(data)

# First 10 values of each variable
head(data)

# Take a look at the data
View(data)

# Check the number of rows (observations) and columns (variables)
nrow(data)

ncol(data)
```
