---
eval: true
code-fold: false
engine: knitr
---

# R basics

In this chapter we will introduce to the R basics and some exercises to get familiar to how R works.

## Simple operations

### Math operations

#### **Sum**

```{r}
1+1
```

#### **Subtraction**

```{r}
5-2
```

#### **Multiplication**

```{r}
2*2
```

#### **Division**

```{r}
8/2
```

#### **Round the number**

```{r}
round(3.14)
round(3.14, 1) # The "1" indicates to round it up to 1 decimal digit.

# You can use help ?round in the console to see the description of the function. 
```

### Basic shortpaths

#### **Perform Combinations**

```{r}
c(1, 2, 3)
c(1:3) # The ":" indicates a range between the first and second numbers. 
```

#### **Create a comment with ctrl + shift + r**

```{r}

# Comments help you organize your code. The software will not run the comment. 
```

#### Create a table

A simple table with the number of trips by car, PT, walking, and cycling in a hypothetical street segment at a certain period.

**Define the variables**

```{r}
modes <- c("car", "PT", "walking", "cycling") # you can use "=" or "<-"
Trips = c(200, 50, 300, 150) # uppercase letters modify

```

**Join the variables to create a table**

```{r}
table_example = data.frame(modes, Trips)
```

**Take a look at the table**

Visualize the table by clicking on the "Data" in the "Environment" page or use the following function.

```{r}
View(table_example)
```

## Practical exercise

**Dataset:** the number of trips between all municipalities in the Metropolitan Area of Lisbon, Portugal [@IMOB].

#### Import dataset

```{r}
data = readRDS("data/TRIPSmode_mun.Rds")
```

#### Take a first look at the data

**Summary statistics**

```{r}
summary(data)
```

**Check the structure of the data**

```{r}
str(data)
```

**Check the first 10 values of each variable**

```{r}
head(data, 10)
```

**Check the number of rows (observations) and columns (variables)**

```{r}
nrow(data)

ncol(data)
```

**Open the dataset**

```{r}
View(data)
```

#### Explore the data

**Check the total number of trips**

Use '\$' to select a variable of the Data

```{r}
sum(data$Total)
```

**Percentage of car trips related to the total**

```{r}
sum(data$Car)/sum(data$Total) * 100
```

**Percentage of active modes related to the total**

```{r}
(sum(data$Walk)+ sum(data$Bike)) / sum(data$Total) * 100
```

#### Modify original data

**Create a column with the sum of the number of trips for active modes**

```{r}
data$Active = data$Walk + data$Bike
```

**Filter by condition (create new tables)**

Filter trips only with origin from Lisbon

```{r}
data_Lisbon = data[data$Origin_mun == "Lisboa",]
```

Filter trips with origin **different** from Lisbon

```{r}
data_out_Lisbon = data[data$Origin_mun != "Lisboa",]
```

Filter trips with origin and destination in Lisbon

```{r}
data_in_Out_Lisbon = data[data$Origin_mun == "Lisboa" & data$Destination_mun == "Lisboa",]
```

#### Modify original data

**Create a column**

The sum of the number of trips for active modes

```{r}
data$Active = data$Walk + data$Bike
```

**Remove a column**

Look at the first row

```{r}
data[1,] #rows and columns start from 1
```

Look at first row and column

```{r}
data[1,1]
```

Remove the first column

```{r}
data = data[,-1] #first column
```

**Create a table only with origin, destination and walking trips**

There are many ways to do the same operation.

```{r}
names(data)
```

```{r}
data_walk2 = data[,c(1,2,4)]
```

```{r}
data_walk3 = data[,-c(3,5:9)]
```

#### Export data

Save data in **.csv** and **.Rds**

```{r}
write.csv(data, 'dataset.csv', row.names = FALSE)
saveRDS(data, 'data/dataset.Rds') #Choose a different file. 
```

#### Import data

```{r}
csv_file = read.csv("dataset.csv")
rds_file =readRDS("data/dataset.Rds")
```
